# (PART\*) Descriptive Statistics {-}


# Wide Fom

## Import Data

We're going to use the `twinData` dataset from the `OpenMx` package. This dataset contains 3,808 pairs of Australian twins on the 12 variables.


```{r libraries}
library(tidyverse)
library(NlsyLinks)
library(discord)
library(BGmisc)
library(OpenMx)

data(twinData)

df <- twinData
```

## Data Structure

```{r data-structure}
class(df)
glimpse(df)
```

```{r addsex}

df <- df %>% mutate(sex = 
                      case_when(zygosity %in% c("MZFF","DZFF") ~ "F", 
                                zygosity %in% c("MZMM","DZMM") ~ "M",
                                TRUE ~ "OS"),
                    zyg = case_when(zygosity %in% c("MZFF","MZMM") ~ "MZ", 
                                    zygosity %in% c("DZFF","DZMM","DZOS") ~ "DZ",
                                TRUE ~ NA_character_ ))
```

## Summary Statistics

### Numeric Variables

```{r summary-stats}

# Calculate summary statistics for numeric variables across the full sample
summary_stats <- df %>%
  summarise(across(where(is.numeric), list(
    mean = ~mean(., na.rm = TRUE),
    sd = ~sd(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    min = ~min(., na.rm = TRUE),
    max = ~max(., na.rm = TRUE),
    IQR = ~IQR(., na.rm = TRUE)
  ), .names = "{col}_{fn}")) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "statistic"),
    names_sep = "_"
  ) %>%
  pivot_wider(
    names_from = statistic,
    values_from = value
  )

summary_stats
```


## Frequency Tables

```{r frequency-tables}

# Counting 'zygosity' and calculating percentages
zygosity_summary <- df %>%
  count(zyg, name = "count") %>%
  mutate(percentage = count / sum(count) * 100) %>%
  rename(category = zyg) %>%  # Renaming the column for clarity
  mutate(variable = "zygosity")  # Adding a descriptor column for the variable

# Counting 'sex' and calculating percentages
sex_summary <- df %>%
  count(sex, name = "count") %>%
  mutate(percentage = count / sum(count) * 100) %>%
  rename(category = sex) %>%  # Renaming the column for clarity
  mutate(variable = "sex")  # Adding a descriptor column for the variable

# Combining both summaries into a single dataframe
combined_summary <- bind_rows(zygosity_summary, sex_summary) %>%
  select(variable, category, everything())  # Reordering columns for clarity

combined_summary

```

What if you want to examine effects by cohort? You can use the `group_by` function to group the data by cohort and then calculate the summary statistics for each cohort.

```{r summary-stats-cohort}

library(tidyverse)

# Grouping by 'cohort' and calculating summary statistics for each group across values that are numeric
df_summary <- df %>%
    group_by(cohort) %>%
    summarise(across(where(is.numeric), list(
        mean = ~mean(., na.rm = TRUE),
        sd = ~sd(., na.rm = TRUE),
        median = ~median(., na.rm = TRUE),
        IQR = ~IQR(., na.rm = TRUE)
    ), .names = "{.col}_{.fn}")) %>%
    # Ensure only statistic columns are being pivoted
    pivot_longer(
        cols = -cohort,  # Exclude 'cohort' from pivoting
        names_to = c("variable", "statistic"),
        names_sep = "_"
    )
# lock in the variable order
variable_order <- unique(df_summary$variable)

df_summary <- df_summary  %>% 
    pivot_wider(
        names_from = statistic,
        values_from = value,
        values_fn = list(value = mean)  # Aggregate if necessary
    )

df_summary  %>% mutate(variable = factor(variable, levels = variable_order)) %>%
    arrange(variable, cohort)

```


What about descriptives by zygosity and sex?

```{r summary-stats-zyg}

library(tidyverse)

# Grouping by 'zyg,sex' and calculating summary statistics for each group
df_summary <- df %>%
    group_by(zyg,sex) %>%
    summarise(across(where(is.numeric), list(
        mean = ~mean(., na.rm = TRUE),
        sd = ~sd(., na.rm = TRUE),
        median = ~median(., na.rm = TRUE),
        IQR = ~IQR(., na.rm = TRUE)
    ), .names = "{.col}_{.fn}")) %>%
    # Ensure only statistic columns are being pivoted
    pivot_longer(
        cols = -c(zyg,sex),  # Exclude 'zygosity' from pivoting
        names_to = c("variable", "statistic"),
        names_sep = "_"
    )
# lock in the variable order
variable_order <- unique(df_summary$variable)

df_summary <- df_summary  %>% 
    pivot_wider(
        names_from = statistic,
        values_from = value,
        values_fn = list(value = mean)  # Aggregate if necessary
    )

df_summary  %>% mutate(variable = factor(variable, levels = variable_order)) %>%
    arrange(variable,zyg,sex)

```

## Plots

Histograms and scatter plots to visualize distributions and relationships.


```{r data-viz, fig.width=8, fig.height=4}
ggplot(df, aes(x = wt1)) +
  geom_histogram(bins=30, fill="blue", color="black") +
  labs(x="weight", y="Frequency", title="Distribution of weight for Twin 1") +
  theme_minimal()
```
```{r}
# Basic Scatter Plot of weight of Twin 1 vs. weight of Twin 2
p <- ggplot(df, aes(x=wt1, y=wt2, color=zyg)) +
  geom_point(alpha=.5) +
  labs(x = "Weight of Twin 1", 
       y = "Weight of Twin 2", 
       title = "Scatterplot of weight by Zygosity") +
  scale_color_viridis_d(option = "virdis",
                        begin = 0.1,end=.85) +
  theme_minimal()

p
```

Adding a regression line to the scatter plot.

```{r}
p + geom_smooth(method = "lm", se = FALSE)

```

```{r}

library(ggplot2)
library(ggExtra)


# Create marginal density plots for x and y axes
p_x <- ggplot(df, aes(x = wt1, fill = zyg)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  scale_fill_viridis_d(option = "viridis", begin = 0.1, end = 0.85) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

p_x


p_y <- ggplot(df, aes(x = wt2, fill = zyg)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis_d(option = "viridis", begin = 0.1, end = 0.85) +
  coord_flip() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

 p_y
 
 
# with marginal histogram
p1 <- ggMarginal(p, type="histogram")
 
p1
```
```{r}
# marginal density
p2 <- ggMarginal(p, type="density")
p2
```

```{r}
# marginal boxplot
p3 <- ggMarginal(p, type="boxplot")
p3
```


## Correlation Matrix

```{r correlation-matrix}

# Calculate the correlation matrix

```
