
---
title: "Visualizing Twin Data with ggplot2"
author: "S. Mason Garrison"
output: html_document
---

# Coding out loud

## Start with the `twinData_cleaned` data frame

.midi[
> **Start with the `twinData_cleaned` data frame**
]

.pull-left[
```{r twin-0, fig.show="hide", warning=FALSE}
ggplot(data = twinData_cleaned)
```
]
.pull-right[
```{r ref.label = "twin-0", echo = FALSE, warning = FALSE, out.width = "100%", fig.width = 8}
```
]

## Map height of Twin 1 to the x-axis

.midi[
> Start with the `twinData_cleaned` data frame,
> **map height of Twin 1 to the x-axis**
]

.pull-left[
```{r twin-1, fig.show = "hide", warning = FALSE}
ggplot(data = twinData_cleaned,
       mapping = aes(x = height_t1))
```
]
.pull-right[
```{r ref.label = "twin-1", echo = FALSE, warning = FALSE, out.width = "100%", fig.width = 8}
```
]

## Map height of Twin 2 to the y-axis

.midi[
> Start with the `twinData_cleaned` data frame,
> map height of Twin 1 to the x-axis
> **and map height of Twin 2 to the y-axis.**
]

.pull-left[
```{r twin-2, fig.show = "hide", warning = FALSE}
ggplot(data = twinData_cleaned,
       mapping = aes(x = height_t1, y = height_t2))
```
]
.pull-right[
```{r ref.label = "twin-2", echo = FALSE, warning = FALSE, out.width = "100%", fig.width 8}
```
]

## Represent each observation with a point

.midi[
> Start with the `twinData_cleaned` data frame,
> map height of Twin 1 to the x-axis
> and map height of Twin 2 to the y-axis. 
> **Represent each observation with a point**
]

.pull-left[
```{r twin-3, fig.show = "hide", warning = FALSE}
ggplot(data = twinData_cleaned,
       mapping = aes(x = height_t1, y = height_t2)) + 
  geom_point()
```
]
.pull-right[
```{r ref.label = "twin-3", echo = FALSE, warning = FALSE, out.width = "100%", fig.width 8}
```
]

## Color by zygosity

.midi[
> Start with the `twinData_cleaned` data frame,
> map height of Twin 1 to the x-axis
> and map height of Twin 2 to the y-axis. 
> Represent each observation with a point
> **and map zygosity to the color of each point.**
]

.pull-left[
```{r twin-4, fig.show = "hide", warning = FALSE}
ggplot(data = twinData_cleaned,
       mapping = aes(x = height_t1, y = height_t2, color = zyg)) +
  geom_point()
```
]
.pull-right[
```{r ref.label = "twin-4", echo = FALSE, warning = FALSE, out.width = "100%", fig.width 8}
```
]
