
## Add titles and labels

.midi[
> Start with the `twinData_cleaned` data frame,
> map height of Twin 1 to the x-axis,
> and map height of Twin 2 to the y-axis.
> Represent each observation with a point,
> and map zygosity to the color of each point.
> **Title the plot 'Height Comparison between Twins',**
> **add subtitles and label axes.**
]

.pull-left[
```{r twin-5, fig.show = "hide", warning=FALSE}
ggplot(data = twinData_cleaned, mapping = aes(x = height_t1, y = height_t2, color = zyg)) +
  geom_point() +
  labs(title = "Height Comparison between Twins", subtitle = "Visualized by Zygosity",
       x = "Height of Twin 1 (cm)", y = "Height of Twin 2 (cm)", color = "Zygosity") #<<
```
]
.pull-right[
```{r ref.label = "twin-5", echo = FALSE, warning = FALSE, out.width = "100%", fig.width 8}
```
]

## Add a caption for the data source

.midi[
> Start with the `twinData_cleaned` data frame,
> map height of Twin 1 to the x-axis,
> and map height of Twin 2 to the y-axis.
> Represent each observation with a point,
> and map zygosity to the color of each point.
> Title the plot "Height Comparison between Twins",
> add the subtitle "Visualized by Zygosity",
> label the x and y axes,
> **and add a caption for the data source.**
]

.pull-left[
```{r twin-6, fig.show = "hide", warning=FALSE}
ggplot(data = twinData_cleaned, mapping = aes(x = height_t1, y = height_t2, color = zyg)) +
  geom_point() +
  labs(title = "Height Comparison between Twins", subtitle = "Visualized by Zygosity",
       x = "Height of Twin 1 (cm)", y = "Height of Twin 2 (cm)", color = "Zygosity",
       caption = "Source: Australian National Health and Medical Research Council Twin Registry") #<<
```
]
.pull-right[
```{r ref.label = "twin-6", echo = FALSE, warning = FALSE, out.width = "100%", fig.width 8}
```
]

## Use a discrete color scale for color blindness

.midi[
> Start with the `twinData_cleaned` data frame,
> map height of Twin 1 to the x-axis,
> and map height of Twin 2 to the y-axis.
> Represent each observation with a point,
> and map zygosity to the color of each point.
> Title the plot "Height Comparison between Twins",
> add the subtitle "Visualized by Zygosity",
> label the x and y axes,
> add a caption for the data source,
> **and use a discrete color scale that is designed to be perceived by viewers with common forms of color blindness.**
]

.pull-left[
```{r twin-7, fig.show = "hide", warning=FALSE}
ggplot(data = twinData_cleaned, mapping = aes(x = height_t1, y = height_t2, color = zyg)) +
  geom_point() +
  labs(title = "Height Comparison between Twins", subtitle = "Visualized by Zygosity",
       x = "Height of Twin 1 (cm)", y = "Height of Twin 2 (cm)", color = "Zygosity",
       caption = "Source: Australian National Health and Medical Research Council Twin Registry") +
  scale_color_viridis_d() #<<
```
]
.pull-right[
```{r ref.label = "twin-7", echo = FALSE, warning = FALSE, out.width = "100%", fig.width 8}
```
]
